### 2015-07-06 chapters/ipa_meets_unicode.tex
adds rich history of IPA and how it was incorporated into Unicode

@@

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into Latin-based characters. Although IPA
is reasonably easily adhered to with pen and paper, it is not trivial to encode
IPA characters electronically. Similar to the previous chapter, this chapter
will discuss various pitfalls with the encoding of IPA in the Unicode Standard.\@ We will
specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
version of Unicode \citep{Unicode2014}.

\noindent For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to those code points. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Further, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a common example of the non-interoperability of
data and data formats.

To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
Methods Phonetic Alphabet) was created to represent IPA symbols with ASCII
character sequences, e.g. <p\textbackslash> for [ɸ]. Two problems with SAMPA are
that (i) it is only a partial encoding of the IPA and (ii) it encodes different
languages in separate data tables, instead of using a universal alphabet, like
IPA.\@ SAMPA tables are derived from phonemes appearing in several European
languages that were developed as part of a European Commission-funded project to
address technical problems like electronic mail exchange (what is now simply
called email). SAMPA is essentially a hack to work around displaying IPA
characters, but it provided speech technology and other fields a basis that has
been widely adopted and used in code. So, SAMPA was a collection of tables to be
compared, instead of a large universal table representing all languages. An
extended version of SAMPA, called X-SAMPA, set out to include every symbol in
the IPA chart including all diacritics \citep{WellsND}. X-SAMPA was considered
more universally applicable because it consisted of one table that encoded all
characters that represented phones/segments in IPA across languages. SAMPA and
X-SAMPA have been widely used for speech technology and as an encoding system in
computational linguistics. Eventually, ASCII-encoding of the IPA became
deprecated through the advent of the Unicode Standard. Note however that many
popular software packages used for linguistic analyses still require ASCII
input, e.g.~RuG/L04 and SplitsTree4.\footnote{See
\url{http://www.let.rug.nl/kleiweg/L04/} and \url{http://www.splitstree.org/},
respectively}

There are several pitfalls to be aware of when using the Unicode Standard to
encode IPA.\@ As we have said before, from a linguistic perspective it might
sometimes look like the Unicode Consortium is making incomprehensible decisions,
but it is important to realize that the consortium has tried and is continuing
to try to be as consistent as possible across a wide range of use cases, and it
does place linguistic traditions above other orthographic possibilities. In
general, we strongly suggest to linguists to not complain about any decisions in
the Unicode Standard, but to try and understand the rationale of the Unicode
Consortium (which in our experience is almost always well-conceived) and devise
ways to work with any unexpected behavior. Many of the current problems derive
from the fact that the IPA is clearly historically based on the Latin script,
but different enough from most other Latin-based writing systems to warrant
special attention. This ambivalent status of the IPA glyphs (partly Latin,
partly special) is unfortunately also attested in the treatment of IPA in the
Unicode Standard. In retrospect, it might have been better to consider the IPA
(and other transcription systems) to be a special kind of script within
the Unicode Standard, and treat the obvious similarity to Latin glyphs as a
historical relic. All IPA glyphs would then have their own code points, instead
of the current situation in which some IPA glyphs have special code points,
while others are treated as being identical to the regular Latin characters.
Yet, the current situation, however unfortunate, is unlikely to change, so as
linguists we will have to learn to deal with the specific pitfalls of IPA within
the Unicode Standard. In this section, we will describe these pitfalls in some
detail.

>>>>

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into discrete segments using a Latin-based 
alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode
IPA characters electronically. Similar to the previous chapter, in this chapter
we discuss various pitfalls with the encoding of IPA in the Unicode Standard.\@ We will
specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
version of Unicode \citep{Unicode2014}.

There are several pitfalls to be aware of when using the Unicode Standard to
encode IPA.\@ As we have said before, from a linguistic perspective it might
sometimes look like the Unicode Consortium is making incomprehensible decisions,
but it is important to realize that the consortium has tried and is continuing
to try to be as consistent as possible across a wide range of use cases, and it
does place linguistic traditions above other orthographic possibilities. Furthermore, 
when we look at the history of how the IPA meet Unicode, we see that many of the 
decisions for IPA symbols in the Unicode Standard come directly from the 
International Phonetic Association itself. Therefore as users, many pitfalls that we 
encounter have their grounding in the historical principles of the IPA, as 
well as the technological considerations involved in creating a single 
multilingual computing encoding.

In general, we strongly suggest to linguists to not complain about any decisions in
the Unicode Standard, but to try and understand the rationale of the International Phonetic Association 
and the Unicode Consortium (which in our experience is almost always well-conceived) and devise
ways to work with any unexpected behavior. Many of the current problems derive
from the fact that the IPA is clearly historically based on the Latin script,
but different enough from most other Latin-based writing systems to warrant
special attention. This ambivalent status of the IPA glyphs (partly Latin,
partly special) is unfortunately also attested in the treatment of IPA in the
Unicode Standard. In retrospect, it might have been better to consider the IPA
(and other transcription systems) to be a special kind of script within
the Unicode Standard, and treat the obvious similarity to Latin glyphs as a
historical relic. All IPA glyphs would then have their own code points, instead
of the current situation in which some IPA glyphs have special code points,
while others are treated as being identical to the regular Latin characters.
Yet, the current situation, however unfortunate, is unlikely to change, so as
linguists we must learn to deal with the specific pitfalls of IPA within
the Unicode Standard.

In the following sections, we will describe these pitfalls 
in detail. But first we present a brief history of IPA, its principles, 
its computational representation and the need for and challenge in creating 
a single multilingual computer encoding, and then how IPA meet Unicode.

@@

\subsubsection*{Brief history}

Established in 1886, the International Phonetic Association (henceforth Association) 
has long maintained a standard alphabet, the International Phonetic Alphabet (IPA),
\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.} 
which is a common standard in linguistics to transcribe sounds of spoken languages. 
It was first published in 1888 as an international system of phonetic transcription 
for oral languages and for pedagogical purposes. The draft of 1887 contained phonetic 
values for English, French and German. Diacritics for length and nasalization were 
already present in this draft and the same symbols are still used today.

Although the original IPA alphabet was a list of symbols with pronunciation examples 
from words in different languages, in 1900 the symbols were organized into chart and 
given phonetic feature labels for consonants (e.g.\ for manner of articulation: `plosives', 
`nasales', `fricatives'; for place of articulation: `bronchials', `laryngeales', `labiales') 
and for vowels (`fermeées', `mi-fermeées', `mi-overtes', `overtes'). Over the last century 
or so, the structure of the chart has changed with increases in phonetic knowledge, and 
thus like notational systems in other scientific disciplines, the IPA reflects facts and 
theories (of phonetic knowledge) that have developed over time. It is natural then that 
the IPA is modified occasionally to accommodate scientific innovations and discoveries. 
(In fact, this is part of the Association's mandate.) These changes are captured in 
the revisions to the IPA.\footnote{For a detailed history see: 
\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there has been several revisions, many minor. For example, 
articulation labels -- what we now call features even though the IPA avoids this term -- 
have changed (e.g.\ `lips', `throat', `rolled'). Phonetic symbol values have changed 
(e.g.\ voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the 
caret diacritic denoting `long and narrow' is longer used). And many symbols have 
been added to reflect contrastive sounds found in the world's very diverse 
phonological systems.

Although IPA began its life as a pedagogical tool, from its earliest days the 
Association aimed to provide ``a separate sign for each distinctive sound; 
that is, for each sound which, being used instead of another, in the same 
language, can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive 
sounds became later known as \textsc{phonemes} and the IPA has developed historically 
into a notational devise with a strictly segmented phonemic view. A phoneme is an 
abstract theoretical notion derived from an acoustic signal as produced by speakers 
in the real world. Therefore the IPA contains a number of theoretical assumptions 
about speech and how to analyze speech in written form.

Phonetic analysis is based on two premises: that it is possible to describe 
the acoustic speech signal (i.e.\ sound waves) in terms of sequentially ordered 
discrete segments, and, that each segment can be characterized by an articulatory 
target.\footnote{A purely phonetic description is only derivable from instrumental 
data in high quality sound recordings. Once spoken language data are segmented, 
phonological consideration inextricably play a roll in transcription. In other 
words, phonetic observations beyond quantitative acoustic analysis are made in 
terms of some phonological framework.} Today, the IPA chart reflects a linguistic 
basis grounded in principles of phonological contrast. This fact is stated 
explicitly in several places, including in the \textit{Report on the 1989 Kiel 
convention} published in the \textit{Journal of the International Phonetic 
Association} \citep[67-68]{International1989report}:

\begin{quote}
The IPA is intended to be a set of symbols for representing all the possible 
sounds of the world's languages. The representation of these sounds uses a set 
of phonetic categories which describe how each sound is made. These categories 
define a number of natural classes of sounds that operate in phonological rules 
and historical sound changes. The symbols of the IPA are shorthand ways of 
indicating certain intersections of these categories.
\end{quote}

\noindent and in the \textit{Handbook of the International Phonetic Association} \citep[18]{IPA1993}:

\begin{quote}
% The general value of the symbols in the chart is listed below. In each case 
... a symbol can be regarded as a shorthand equivalent to a phonetic description, 
and a way of representing the contrasting sounds that occur in a language. Thus 
[m] is equivalent to `voiced bilabial nasal', and is also a way of representing 
one of the contrasting nasal sounds that occur in English and other languages. 
[...] When a symbol is said to be suitable for the representation of sounds in 
two languages, it does not necessarily mean that the sounds in the two languages 
are identical.
\end{quote}

\noindent Although the IPA provides symbols to unambiguously represent phonemes, 
it also aims to represent phonetic details. Since phonetic detail could potentially 
include things like `deep voice', the IPA restricts phonetic detail to linguistically 
relevant aspects of speech. These principles and conventions for using the IPA are 
outlined in the Association's Handbook.

\subsubsection*{Principles}

IPA transcription has essentially two parts. The first is a text containing IPA 
symbols and the second is a set of conventions (rules) for interpreting those 
symbols (and their combinations). The IPA is designed to meet practical linguistic 
needs and is used to transcribe the phonetic or phonological structure of languages. 
It is also used increasingly as a foreign language learning tool, as a standard 
pronunciation guide and as a tool for creating practical orthographies of 
previously unwritten languages.

The current construction and use of the IPA are guided by principles outlined 
in the \textit{Handbook of the International Phonetic Association: A guide 
to the use of the International Phonetic Alphabet} \citep[159]{IPA1999}, henceforth 
HB. The use of the symbols to represent a language's phonological system is guided 
by the principle of contrast; where two words are distinguishable by phonetic contrast, 
those contrasts should be transcribed with different symbols (graphemes not diacritics). 
Allophonic distinction falls under the rubric of diacritically-distinguished symbols, 
e.g. [stop̚] vs [spʰot]. In sum:

\begin{itemize}
	\item different symbols (without diacritics) should be used whenever a language employs two contrastive sounds
	\item when two sounds in a language are not known to be contrastive, the same symbol should be used to represent these sounds; however, diacritics may be used to distinguish such sounds when necessary
	\item diacritics cannot be dispensed with entirely, so the Association recommends to limit their use to:
	\begin{itemize}
		\item denoting length, stress and pitch
		\item representing minute shades of sounds
		\item obviating the design of a (large) number of new symbols when a single diacritic suffices (e.g. nasalized vowels, aspirated stops)
	\end{itemize}
\end{itemize}

\noindent Thus, an IPA transcription \textit{always} consists of ``a set of 
symbols and a set of conventions for their interpretation''.

In systematic transcription (as opposed to impressionistic), there is a division 
between phonemic and allophonic transcription. The terms phoneme and allophone 
contain theoretical baggage, but the basic goal of a phonemic transcription is to 
distinguish all words in a language with the minimal number of transcription 
symbols \citep[19]{Abercrombie1964}. Allophonic transcription uses a broader 
set of distinct symbols to describe systematic allophonic differences in sounds 
in words. These two systematic transcriptions are related to each other by a 
set of conventions in the IPA tradition, so that they can be converted between 
one and another.

An IPA transcription is connected to a speech event by a set of conventions. 
A phonetic (or impressionistic) transcription may use the conventions implicit 
in the IPA chart, i.e. the transcriber can indicate that the phonetic value of 
<ŋ͡m> is a simultaneous labial and velar closure which is voiced and contains 
nasal airflow.

A phonemic transcription includes the conventions of a particular language's 
phonological rules. These rules determine the realization of that language's 
phonemes. However, there can be different systems of phonemic transcription 
for the same variety of a language. The differences may result from the fact 
that more than one phonetic symbol may be appropriate for a phoneme (see Section 
\ref{}). Or the differences may be due to different phonemic analyses, e.g.\ 
Standard German's vowel system is arguably contrastive in length or tenseness.

An important principle of the IPA is that different representations resulting 
from different symbols and different analyses are in line with the IPA's aims. 
In other words, the IPA does not provide phonological analyses for specific 
languages and the IPA does not define a single ``correct'' transcription system. 
Rather, the IPA aims to provide a resource that allows users to express any 
analysis so that it is widely understood. Thus the IPA suits many linguists' 
needs because:

\begin{itemize}
	\item it is intended to be a set of symbols for representing all possible sounds in the world's (spoken) languages
	\item its chart has a linguistic basis (specifically a phonological bias) rather than a general phonetic notation scheme
	\item its symbols can be used to represent distinctive feature combinations\footnote{Although the chart uses traditional manner and place of articulation labels, the symbols can nevertheless represent any defined bundle of features, binary or otherwise, to define phonetic dimensions.}
	\item its chart provides a summary of linguists' agreed-upon phonetic knowledge (a common denominator of phonological ``facts'')
\end{itemize}

Several styles of transcription with IPA are possible and the HB illustrates 
these and notes that they are all valid (see the 29 languages and their transcriptions 
in the original and initial \textit{Illustrations of the IPA} \citep[41--154]{IPA2007}). 
Therefore, there are different but equivalent transcriptions, or as \cite[64]{Ladefoged1990a} 
captures it, ``Perhaps now that the Association has been explicit in its eclectic 
approach, outsiders to the Association will no longer speak of \textit{the} IPA 
transcription of a given phenomenon, as if there were only one approved style.''.

Clearly not all phoneticians agree (or will likely ever agree of course) on all 
aspects of the IPA or on transcription practices. As noted above, there have 
been several revisions in the IPA's long history, but the current version (2005) 
is strikingly similar to the 1926 version. In 1989 an IPA revision convention 
was held in Kiel, Germany. As per other revisions, there was expansion and revisions 
to phonetic symbols in the IPA chart. Notably the marking of tone was extended 
with the addition of a second system for marking linguistic tones (Chao tones). 
Importantly, however, this was the first revision to address issues of 
computational representations for the IPA -- the principles above of which 
have had several ramifications that make the interoperability of electronic 
linguistic data extremely difficult.

\subsubsection*{Computational representation}

Prior to the Kiel Convention for the modern revision of the IPA in 1989, 
\cite{Wells1987} collected and published practical approaches to coded 
representations of the IPA, which dealt mainly with the assignment of 
characters on the keyboard. The process of assigning standardized ``computer 
codes'' to phonetic symbols was assigned to the Workgroup on Computer 
Coding (henceforth working group) at the Kiel Convention. This working 
group was tasked with \citep{Esling1990,EslingGaylord1993}:

\begin{itemize}
	\item determining how to represent the IPA numerically
	\item developing a set of numbers to refer to the IPA symbols unambiguously
	\item providing each symbol a unique name (intended to provide a mnemonic description of that character's shape)
\end{itemize}

\noindent The identification of IPA symbols with unique identifiers was 
a first step in formalizing the IPA computationally because it would give 
each symbol an unambiguous numerical identifier called an \textsc{IPA Number}. 
The numbering system was to be comprehensive enough to support future revisions 
of the IPA, including symbol specifications and diacritic placement. The 
application of diacritics was also to be made explicit.

Although the Association had never officially approved a set of names 
for the IPA symbols, each IPA symbol received a unique \textsc{IPA Name}. 
Many symbols already had an informal name (or two) used by linguists, but 
consensus on symbol names was growing due to the recent publication of the 
\textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
IPA symbol names were taken from \cite{PullumLadusaw1986} \citep[31]{IPA1999}.

The working group insightfully decided that the computing-coding convention 
for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
i.e.\ the IPA Number was not meant to be implemented directly in a computer 
encoding. The working group report's declaration includes the explanatory 
remarks \citep[82]{International1989report}:

\begin{quote}
The recommendation of a 7-bit ASCII or 8-bit extended-ASCII coding system 
would be short-sighted in view of development towards 16-bit and 32-bit 
processors. In fact, any specific recommendations would tie the Association 
to a stage of technological development which is bound to be outdated long 
before the next revision of the handbook.
\end{quote}

\noindent Thus the coding convention was not meant to address the engineering 
aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
it was meant to serve as a basis for a communication-interchange standard for 
creating mapping tables from various computer encodings, fonts, phonetic-character-set 
software, etc., to common IPA Numbers, and thus symbols.\footnote{Remember, at 
this time in the late 1980s there was no stable multilingual computing environment. 
But some solution was needed because scholars were increasingly using personal 
computers for their research and many were quickly adopting electronic mail or 
discussion boards like Usenet as a medium for international exchanges. This was 
before the Internet as we know it today. Most of these systems ran on 8-bit 
hardware systems using a 7-bit ASCII character encoding.}

Furthermore, the assignment of computer codes to IPA symbols was meant to 
represent an unbiased formulation. The Association plays the role of an international 
advisory body and it stated that it should not recommend a particular existing 
system of encoding. In fact, during this time there were a number of coding 
systems used, but none of them had a dominant international position. The 
differences between systems were also either too great or too subtle to warrant 
an attempt at combining them \citep{International1989report}.

The working group assigned each IPA symbol to a unique three-digit number, 
i.e.\ an IPA Number. Encoded in this number scheme is information about the 
status of each symbol (see below). The IPA numbers are listed with the IPA 
symbols and they are also illustrated in IPA chart form (see \cite[84]{EslingGaylord1993} 
or \cite[App. 2]{IPA2007}). The numbers were assigned in linear order (e.g.\ 
[p] 101, [b] 102, [t] 103...) following the IPA revision of 1989 and its update in 1996.

The working group made the decision that no IPA symbol, past or present, 
could be ignored. The comprehensive inclusion of all IPA symbols was to 
anticipate the possibility that some symbols might be added, withdrawn, 
or reintroduced into current or future usage. For example, in the 1989 
revision voiceless implosives < ƥ, ƭ, ƈ, ƙ, ʠ > were added; in the 1993 
revision they were removed. Ligatures like < ʧ, ʤ > are included as formerly 
recognized IPA symbols; they are assigned to the ``200 series'' of IPA numbers 
as members of the group of symbols formerly recognized by the IPA. To ensure 
backwards compatibility, legacy IPA symbols would retain an IPA Number and 
an IPA Name for reference purposes. As we discuss below, this decision is 
later reflected in the Unicode Standard; many legacy IPA symbols reside in 
the \textsc{IPA Extensions} block.

The IPA Number is simply expressed as a ``three-digit number numerical 
directory of digit triples'' \cite{}.\footnote{For practical purposes, 
the IPA Number also served as a typesetter's guide to the IPA chart.} 
The numbering scheme specifies three-digit codes, the first digit of which 
indicates the symbol's category \cite{Esling1990,EslingGaylord1993}:

\begin{itemize}
	\item 100s for accepted IPA consonants
	\item 200s for former IPA consonants and non-IPA symbols
	\item 300s for vowels
	\item 400s for segmental diacritics
	\item 500s for suprasegmental symbols
	\item 600s-800s for future specifications
	\item 900s for escape sequences
\end{itemize}

After a symbol is categorized, it is assigned a number sequentially, 
e.g.\ [i] 301, [e] 302, [ɛ] 303. The system allows for the addition 
of new symbols within the various series by appending them, e.g. [ⱱ] 
184. Former or often used but non-IPA symbols for consonants, vowels 
and diacritics are numbered from x99 backwards. For example, the voiceless 
and voiced postalveolar affricates and fricatives < č, ǰ, š, ž > are 
assigned the IPA numbers 299, 298, 297 and 296, respectively, because 
they are not sanctioned IPA symbols.

The assignment of the IPA numbers to IPA symbols provided the basis for 
uniquely identifying the set of past and present IPA symbols as a type of 
computational representational standard of the IPA. Within each revision 
of the IPA, the coding defines a closed and clearly defined set of characters. 
The benefits of this standardization are clear in at least two ways: it is 
used in translation tables that reference ASCII representations of the IPA, 
and this early computational representation of the IPA became the basis for 
its inclusion into the Unicode Standard version 1.0.

\subsubsection*{SAMPA and X-SAMPA}

True to the working group's aim, the IPA numbers provided a mechanism for 
a communication interchange standard for creating mapping tables to various 
computer encodings. For example, the IPA coding system was used as a mapping 
system in the creation of SAMPA \citep{Wells_etal1992}, an ASCII representation 
of the IPA symbols.

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences as IPA symbol codings.\footnote{\cite{Wells1987} 
provides an in-depth description of IPA codings from country-to-country. 
Later ASCII-IPAs include Kirshenbaum (created in 1992 in a Usenet group and 
named after its lead developer who was at Hewlett-Packard Laboratories) and 
Worldbet (published in 1993 by \cite{Hieronymus1993}, who was at AT\&T Laboratories).} 
A successful effort was SAMPA (Speech Assessment
Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
represent IPA symbols with ASCII
character sequences \citep{Wells1987,Wells_etal1992}, e.g. <p\textbackslash> 
for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
of the principal European Union languages \citep{Wells1995}. It is still widely 
used in language technology.

Two problems with SAMPA are that (i) it is only a partial encoding of the IPA and (ii) it encodes different
languages in separate data tables, instead of using a universal alphabet, like
IPA.\@ SAMPA tables were developed as part of a European Commission-funded project to
address technical problems like electronic mail exchange (what is now simply
called email). SAMPA is essentially a hack to work around displaying IPA
characters, but it provided speech technology and other fields a basis that has
been widely adopted and often still used in code. So, SAMPA is a collection of tables to be
compared, instead of a large universal table representing all languages.

An extended version of SAMPA, called X-SAMPA, set out to include every symbol, 
including all diacritics, in the IPA chart \citep{Wells1995}. X-SAMPA is considered
more universally applicable because it consists of one table that encodes all
characters in IPA. In line with the principles of the IPA, SAMPA and X-SAMPA include a 
repertoire of symbols. These symbols are intended to represent phonemes rather than 
all allophonic distinctions. Additionally, both ASCII-ifications of IPA are useful because 
strings of SAMPA or X-SAMPA are (reportedly) uniquely parsable \citep{Wells1995}. However, 
like the IPA, X-SAMPA has different notations for encoding the same phonetic phenomena 
(see Pitfall \ref{}).

SAMPA and X-SAMPA have been widely used for speech technology and as an encoding system in
computational linguistics. In fact, they are stilled used in popular software packages 
that require ASCII input and some of which have been co-opted for linguistic analyses, 
e.g.~RuG/L04 and SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and 
\url{http://www.splitstree.org/}, respectively}

\subsubsection*{The need for a single multilingual environment}

During the 1980's, it became increasingly clear that an adequate solution 
to the problem of multilingual computing environments was needed. Linguists 
were on the forefront of addressing this issue because they faced these 
challenges head-on by wishing to publish and communicate electronic text 
with phonetic symbols which were not included in basic ASCII.\footnote{One 
only needs to look at facsimiles of older electronic documents to see exotic 
symbols written in by hand.}

Long familiar were linguists already with the distinction between function 
and form. Even in the context of the computer implementation of writing systems, 
the necessity to distinguish form and function had been made \citep{Becker1984}. 
The computer industry, on the other hand, did not consider, ignored, or simply 
did not encode this principle when creating operating systems like MS-DOS, which 
were limited to 256 code points (due to computer hardware architecture) and 
encoded with one-to-one mappings from character codes to graphemes.

Industry was starting to tackle the issues involved in developing a single 
multilingual computing environment on a variety of fronts, including the then 
new technology of bitmap fonts and the creation of Font Manager and Script 
Manager by Apple \citep{Apple1985,Apple1986,Apple1988}. As noted above, around 
this time linguists were developing work-arounds such as SAMPA, so that they 
could communicate IPA transcription and use ASCII-based software. Some linguists 
formalized the issues of multilingual text processing from a computational 
perspective \citep{Anderson1984,Becker1984,Simons1989}. The study of writing 
systems was also being invigorated by the computational challenges in making 
computers work in a multilingual environment.\footnote{\cite[11--15]{Sampson1985} 
urges linguists to view the study of writing systems as legitimate scientific enquiry.}

% This standard became the basis for a proposal to include the IPA in the first version of the Unicode Standard. Decisions by the Computer Coding working group and work they continued after the 1989 Kiel Convention were adopted by the International Phonetic Association. These decisions are directly reflected in the Unicode Standard's encoding of IPA, seeing as it was the Association who submitted the script proposal to the Unicode Consortium.

The second major benefit of the standardization of the IPA in a computational 
representation by the Kiel working group is that it provided the basis for a 
formal proposal to be submitted to various international standards organizations, 
several of which were trying to tackle (and in a sense `win') the multilingual 
computing environment problem. Basically, everyone from corporations to governments 
to language scientists (for lack of a better term) wanted a single unified multilingual 
character encoding set for all the world's writing systems, even if they did not 
understand or appreciate the challenges involved in creating and adopting a solution. 
Additionally, advancements in computer hardware were making the solutions easier 
to implement in software.

The engineering problems and solutions had been spelled out years before, e.g.\ 
a two-byte encoding for multilingual text \citep{Anderson1984}. 
Although languages vary to an astounding extent (cf.\ \cite{EvansLevinson2009}), 
writing systems are quite similar formally and the issue of formal representation 
of the world's orthographic systems had also been addressed \citep{Simons1989}.

A major obstacle in creating a single encoding multilingual environment from 
the perspective of writing systems involves the distinction between function 
and form \citep{Becker1984} This distinction is so central to basic linguistic 
theory and that trained linguists and semiologists take it as second nature. 
A central challenge in developing a universal character set was to combine a 
technological solution with a formalization of writing systems proper.\footnote{Of 
course there were additional practical issues to overcome, e.g.\ funding, creating 
the formal and technological proposal, deciding which characters and writing systems 
to include initially, while setting precedence of how to add new ones in the future.}

In hindsight it is easy to lose sight of how impactful 30 years of technological 
development have been on linguistics, from theory development using quantitative 
means to pure data collection and dissemination (which as fed back into the former). 
But at the end of the 1970s, virtually no ordinary working linguist was using a 
person computer \cite{Simons1996}. Personal computer usage, however, dramatically 
increased throughout the 1980s. By 1990, dozens of character sets were in common use. 
They varied in their architecture and in their character repertoires, which 
made things a mess. There were two major players in the universal character 
set race: Unicode and the International Organization for Standardization (ISO).

\subsubsection*{Unicode and ISO 10646}

In the late 1980s, a universal character set was being developed by what 
is now referred to as the Unicode Consortium.\footnote{The Unicode 
Consortium was officially incorporated in January 1991.} This consortium 
consisted largely, although not entirely, of major US corporations, with 
the aim of overcoming the inoperability of different coded character sets 
and their costly hinderance for developing multilingual software development 
and for internationalization efforts. Commercial importance of course drove 
the early inclusion of Latin, non-Latin, and some exotic scripts; see the 
table of commercial importance as measured by GDP of countries using certain 
writing systems \citep[2]{unicode88}.

The original Unicode manifesto is \cite{unicode88.pdf}.\footnote{http://www.unicode.org/history/unicode88.pdf} 
Its aim was for a reliable international multilingual text encoding standard 
that would encompass all scripts of the world, or in the author's own words, 
``a new, world-wide ASCII''. An in-depth history of Unicode, highlighting 
interesting facts like its first text prototypes at Apple and its incorporation 
into TrueType, is retold online.\footnote{\url{http://www.unicode.org/history/earlyyears.html}}

Unicode 88 provided the basic principles for the Unicode Standard's design -- 
pushing for 16 bit representations of characters with a clear distinction 
between characters and glyphs. Some of the contents of this status proposal 
of 1988 were reworked for inclusion in the early Unicode Standard pre-publication 
drafts and by August 1990, the proposal was in a (very) rough draft format. Its 
editors and the Unicode Working Group (the predecessor to the Unicode Technical 
Committee) worked together to lay out the the proposed standard's structure and 
content. At this time, the proposal contained no code charts nor block descriptions.

% http://www.unicode.org/history/earlyyears.html
% During this period of time, in addition to his co-authoring of Apple KanjiTalk, Davis was involved in further discussions of a universal character set which were being prompted by the development of the Apple File Exchange.

The other major player in developing a universal character set was the ISO 
working group from the International Standards Organization (ISO), based 
in Europe, which was responsible for ISO/IEC 10646. This character set 
standard was composed in 1989 and a draft was published in 1990.\footnote{\url{http://www.iso.org/iso/catalogue_detail.htm?csnumber=56921}} 
The `Universal Multiple-Octet Coded Character Set' or simply UCS was the first 
officially standardized character encoding with the aim of including all characters from all writing systems.\footnote{\url{http://www.nada.kth.se/i18n/ucs/unicode-iso10646-oview.html}}

ISO/IEC 10646 is partly based on ISO/IEC 8859, a series of of ASCII-based 
standard character encodings published in 1987 that use a single bit 8-byte 
character set. Each part of the standard, e.g.\ 8859-1, 8859-5, 8859-6, 
encodes characters to support different languages' writing systems, e.g.\ 
Latin-1 Western European, Latin/Cyrillic, Latin/Arabic, respectively. Being 
a joint effort by the International Organization for Standardization (ISO) 
and the International Electrotechnical Commission (IEC), the aim of the 
standard is reliable information exchange. So, again, issues of phonetic 
symbol encoding, typography, etc., were ignored -- or perhaps more properly 
put, not commercially driven at this early stage.

Intended for the major Western European languages, ISO/IEC 8859 was an 
extension of the ASCII character encoding standard, which included the 
English alphabet, numerals and computer control characters (e.g.\ beep, 
space, carriage return). By extending ASCII's 7-bit system to 8-bit, the 
character repertoire of each of ISO/IEC 8859 character set was doubled 
from 128 to 256 characters. Each character set defined a mapping between 
digital bit patterns and characters, which are visually rendered on screen 
as graphic symbols. ASCII was shared between ISO/IEC 8859 character sets, 
but the characters in the extra bit patterns were different. Thus an aim 
of the ISO working group responsible for ISO/IEC 10646 was to bring all 
characters in all writings systems into a single unified encoding.

In 1991, the Unicode Consortium and the ISO Working Group for ISO/IEC 10646 
decided to create a single universal standard for encoding multilingual text.\footnote{http://unicode.org/book/appC.pdf} 
The two character sets converged, resulting in mutually acceptable changes 
to both, and each group keeps versions of their respective character codes 
and encoding forms synchronized.\footnote{http://www.unicode.org/versions/} 
Although each standard has its own form of reference and the terminology in 
each may differ slightly, the practical difference is that the Unicode Standard 
is a formal implementation of ISO/IEC 10646 and imposes additional constraints 
on its implementation. The Unicode Standard includes character data, algorithms 
and specifications, outside the scope of ISO/IEC 10646, which ensure, when 
properly implemented in software applications and platforms, that characters 
are treated uniformly.

The incorporation of the Unicode Standard into the international encoding 
standard ISO 10646 was approved by ISO as an International Standard in June 
1992.\footnote{http://www.unicode.org/versions/Unicode1.0.0/Notice.pdf} 
The joint Unicode and ISO/IEC 10646 standard has become \textit{the} universal 
character set and it is a single multilingual environment for the majority 
of the world's written languages. Its formal implementation has also been 
vital to the rise of a multi-lingual Internet.

\subsubsection*{IPA and Unicode}

It was a long journey, but the goal of achieving a single multilingual 
computing environment has largely been accomplished. We users, however, 
must cope with the pitfalls that were dug along the way (see Pitfall sections below). 
Some linguists, including your humble authors, are particularly sensitive 
to these issues. So we provide practical advice and approaches in the rest 
of this chapter. But first, we explain how the IPA became incorporated 
into the Unicode Standard via ISO/IEC 10646.

% "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl). The name of each entity is formed by 'IPA' preceding the number, e.g. IPA304 is the rEIentity name of lower-case A. These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text. The advantage of the SGML entity set is that it is independent or the character set being used."
% "A TEl writing system declaration (wsd) has been drawn up for the IPA symbols."
% A TEl writing system declaration (wsd) has been drawn up for the IPA symbols. This document gives information about the symbol and its IPA function, as well as its encoding in the accompanying SGML document and in UnicodelUCS and in AFII. The writing system declaration can be read as a text d9cument or processed by machines in an SMGL process.
